Historical Supply of Biomass from Timber Harvest Residue in West-central Alberta, Canada

### 2021-10-13 analysis/manuscript.Rmd
rolling, rolling

---
title: "Historical Supply of Residual Biomass from Timber Harvest in West-central Alberta, Canada"
author:
- Glen W. Armstrong^[Department of Renewable Resources, University of Alberta]
- Jay A. Anderson^[Department of Resource Economics and Environmental Sociology, University
  of Alberta]
- Alex W. McPhee^[Department of Resource Economics and Environmental Sociology, University
  of Alberta]
date: "`r format(Sys.time(), '%Y-%m-%d')`"
#date: "2021-10-05"
output:
  pdf_document:
    number_sections: yes
    latex_engine: xelatex
    keep_tex: yes
  word_document:
  html_document:
    df_print: paged
csl: canadian-journal-of-forest-research
#csl: biomass-and-bioenergy
bibliography: historicalResidueSupply.bib
---

@@

```{r libraries, echo=FALSE, message=FALSE}
#setwd("/home/gwa/Work/historical-residue/Analysis")
setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp")
#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
library(tidyverse)
library(ggthemes)
library(gridExtra)
library(viridis)
library(skimr)
library(scales)
library(DescTools)
library(zoo)
library(ezknitr)
#ezknit(file="analysis/manuscript.Rmd", out_dir="output", fig_dir="myfigs")
```

@@

Note that the figures are not fixed yet.  I think the "supply curves" are particularly
ugly, but they show our results the way they are.  The choppiness reflects the five minute
service areas, and is valuable because of that.  I am thinking that putting an
average of the 31 years in a different colour in the middle of the cloud might be helpful.

But let's not worry about the details of figure presentation yet, and focus on the writing.

## Some obvious citations

@R2021

@Wickham2019

@qgis2021

@Onyango2019

@Carriquiry2011

@Miles2009

@Dessbesell2017

@Dymond2010

@Lambert2005

@Lo2021

@Malladi2018

@Mansuy2017

@Mansuy2020

@Schroder2019

@Shabani2013

@Shooshtarian2018

@Yemshanov2018

@Zheng2021

@Ackom2010

@Acuna2019

@Boudewyn2007

@Peltola2011

@@

# Materials and methods

Canada's National Forest Inventory System (NFIS) is a product of the Canadian
Council of Forest Ministers (CCFM).  One of the data sets provided by the NFIS
identifies locations of forest harvesting across Canada for the years 1985-2015
detected from Landsat satellite imagery.  These data are at a 30 m spatial 
resolution.  The data were downloaded from
https://opendata.nfis.org/downloads/forest_change/CA_forest_harvest_mask_year_1985_2015.zip
(accessed 2020-03-05).
The data set is described in @Hermosilla2016.

The data were clipped to the Alberta provincial boundary which was obtained from
Altalis (https://www.altalis.com, accessed 2021-09-10), which manages much of the spatial information
belonging to the Alberta government.

## Alberta township coverage

For our analysis, most of our information was aggregated to the township level.
In Alberta, the legal description of any parcel of land uses the Alberta Township
Survey System (see https://www.alberta.ca/alberta-township-survey-system.aspx, 
accessed 2021-10-08). Most townships are roughly square in shape, with an average 
area of about 9,533 ha.  The GIS coverage for township outlines was obtained 
from Altalis (https://www.altalis.com, accessed 2021-09-10).

The harvest areas obtained from the NFIS forest harvesting data set were
aggregated to the township level so that the harvest area for each township
in each of the years from 1985-2015 was recorded in a data set.

## Hypothetical plant locations and capacities

For this study, we assumed that forest harvest residue processing would take place
at a plant co-located with an existing forest products manufacturing facility.
We used the Weyerhaeuser oriented strand board plant in Edson, the Millar Western
Forest Products Ltd. pulp mill in Whitecourt, and the West Fraser pulp mill 
in Hinton as the locations
for the hypothetical forest residue processing facilities.

As well as considering locations for the hypothetical plants, we considered
3 different pland capacities: 200,000 odmt/yr corresponding to a small 
wood pellet mill, 400,000 odmt/yr corresponding to a large pellet mill or
a small ethanol plant, and 800,000 odmt/yr corresponding to a large ethanol 
plant.

## Alberta road network

** A nice picture demonstrating dendritic nature of road network **

  * The provincial road network was sorted into four classes:

    * primary provincial highways

    * other paved roads

    * two-lane gravel roads

    * one-lane gravel or dirt roads

  * Average speeds calculated using harmonic mean:

       $s_h= \left(\frac{s_u^{-1} + s_l^{-1}}{2}\right)^{-1}$

  * Speeds by road class, loaded and unloaded:

    * primary provincial highways: loaded 80, unloaded 95, h-mean 85

    * other paved roads: loaded 65, unloaded 80, h-mean 70

    * two-lane gravel roads: loaded 50, unloaded 70, h-mean 60

    * one-lane roads: loaded 20, unloaded 30, h-mean 25

## service areas

  * Rasterized network areas were generated using the QGIS plugin, QNEAT3 (Iso-Area as Interpolation).

  * QGIS zonal statistics were used to calculate the median network travel time in each township.

## Harvestable volume by township

* based on NRCan 250 m data
    * filtered merch vol >= 50
    * Larches + birches <= 50
    * age >= 80
    * adjusted volumes to give average harvested volumes on West Fraser (Hinton).
    Gives some locally adjusted volumes.  Kludgy, but maybe better than a
    provincial average.

# Theory/Calculation

So here we set the constants.

@@

Following @Roser2013, the pre-piling cost was set to \$`r prepiling_cost`/odmt,
the comminution (*grinding ?*) cost was set to  \$`r comminution_cost`/odmt, and
other costs were set to  \$`r other_cost`/odmt.  The total "constant" costs are 
 \$`r constant_cost`\$/odmt.

Using *blah blah blah* , the truck rate was set to 
\$`r truck_rate`/odmt/h.

A 100 m³ trailer can haul `r odt_trailer` odmt of chipped residue.

@@

harv <- subset(t, harvest_ha > 0)
head(harv)
summary(harv$harvest_ha)
summary(harv$harvest_vol) 
summary(harv$adj_vol)

@@

summary(harv2$harvest_vol)
harv2

harv3 <- read.csv(file = '../data/ABHarvestVol.csv') %>% filter(year >=1990)
harv3

@@

ggplot(data=harv4, aes(x=year)) +
  geom_line(aes(y = harvest_vol), colour="darkred") +
  geom_line(aes(y = harvvol), colour="steelblue") +
  labs(x = "year") + labs(y = "harvest volume (million m3)")  +
  scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6))

#2009 is the median harvest volume year for Alberta in 1990-2015

ggplot(data=harv2, aes(x=year,y=harvest_vol)) +
  labs(x = "year") +
  labs(y = "harvest volume (million m3)") +
  geom_line()  +
  scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6)) 
# +
#   scale_colour_viridis(discrete = TRUE) +
#   theme(legend.position = "none") +
#   facet_grid(cols=vars(location)) +
#   scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))

@@

head(d)

big_joined <- inner_join(d,harv, by = "TRM")
head(big_joined)
tail(big_joined)

@@

```
As a check, we obtained harvest levels for the period 1990-2018 from the National
Forestry Database @ccfm2020 .  Note that in 1998, industrial salvage volumes of 369,791 m³ of 
softwood and 264,647 m³ are not included in the harvest levels shown in the graph

```{r pressure, echo=FALSE, fig.cap="NFDP harvest levels", out.width = '100%'}
knitr::include_graphics("../img/alberta_harvest.png")
```

@@

# Literature cited

>>>>

---
title: Historical Supply of Biomass from Timber Harvest Residue in West-central Alberta, Canada
author:
  - Glen W. Armstrong:
      institute:
        - rr
      correspondence: "yes"
      email: gwa@ualberta.ca
  - Jay A. Anderson:
      email: jaya@ualberta.ca
      institute:
        - rees
  - Alex W. McPhee:
      email: awmcphee@ualberta.ca
      institute: 
        - rees
institute:
  - rr : Department of Renewable Resources, University of Alberta
  - rees: Department of Resource Economics and Environmental Sociology, University of Alberta
date: "`r format(Sys.time(), '%Y-%m-%d')`"
#date: "2021-10-05"
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
output:
  html_document:
    df_print: paged
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
  pdf_document:
    number_sections: yes
    latex_engine: xelatex
    keep_tex: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
  word_document:
     pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
#csl: canadian-journal-of-forest-research
csl: biomass-and-bioenergy
bibliography: historicalResidueSupply.bib
---

<!-- \fontfamily{cmr} -->
<!-- \fontsize{12}{22} -->
<!-- \fontseries{b} -->
<!-- \selectfont -->

@@

```{r libraries, echo=FALSE, message=FALSE}
setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp")
#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
library(tidyverse)
library(ggthemes)
library(gridExtra)
library(viridis)
library(skimr)
library(scales)
library(DescTools)
library(zoo)
library(ezknitr)
#ezknit(file="analysis/manuscript.Rmd", out_dir="output", fig_dir="myfigs")
```

@@

This is **really fucking bold**.

<!-- ## Some obvious citations -->

<!-- @R2021 -->

<!-- @Wickham2019 -->

<!-- @qgis2021 -->

<!-- @Onyango2019 -->

<!-- @Carriquiry2011 -->

<!-- @Miles2009 -->

<!-- @Dessbesell2017 -->

<!-- @Dymond2010 -->

<!-- @Lambert2005 -->

<!-- @Lo2021 -->

<!-- @Malladi2018 -->

<!-- @Mansuy2017 -->

<!-- @Mansuy2020 -->

<!-- @Schroder2019 -->

<!-- @Shabani2013 -->

<!-- @Shooshtarian2018 -->

<!-- @Yemshanov2018 -->

<!-- @Zheng2021 -->

<!-- @Ackom2010 -->

<!-- @Acuna2019 -->

<!-- @Boudewyn2007 -->

<!-- @Peltola2011 -->

@@

@@

```
# Materials and methods

Canada's National Forest Inventory System (NFIS) is a product of the Canadian
Council of Forest Ministers (CCFM).  One of the data sets provided by the NFIS
identifies locations of forest harvesting across Canada for the years 1985-2015
detected from Landsat satellite imagery.  These data are at a 30 m spatial
resolution.  The data were downloaded from
https://opendata.nfis.org/downloads/forest_change/CA_forest_harvest_mask_year_1985_2015.zip
(accessed 2020-03-05). The data set is described in Hermosilla *et al.* (2016)
@Hermosilla2016.
The data were clipped to the Alberta provincial boundary which was obtained from
Altalis (https://www.altalis.com, accessed 2021-09-10), which manages much of
the spatial information belonging to the Alberta government.

## Alberta township coverage
For our analysis, most of our information was aggregated to the township level.
In Alberta, the legal description of any parcel of land uses the Alberta
Township Survey System (see
https://www.alberta.ca/alberta-township-survey-system.aspx, accessed
2021-10-08). Most townships are roughly square in shape, with an average area of
about 9,533 ha.  The GIS coverage for township outlines was obtained from
Altalis (https://www.altalis.com, accessed 2021-09-10).
The harvest areas obtained from the NFIS forest harvesting data set were
aggregated to the township level so that the harvest area for each township in
each of the years from 1985 to 2015 was recorded in a data set.

## Hypothetical plant locations and capacities
For this study, we assumed that forest harvest residue processing would take
place at a plant co-located with an existing forest products manufacturing
facility. We used the Weyerhaeuser oriented strand board plant in Edson, the
Millar Western Forest Products Ltd. pulp mill in Whitecourt, and the West Fraser
pulp mill in Hinton as the locations for the hypothetical forest residue
processing facilities.

As well as considering locations for the hypothetical plants, we considered 3
different plant capacities: 200,000 odmt/yr corresponding to a small wood pellet
mill, 400,000 odmt/yr corresponding to a large pellet mill or a small ethanol
plant, and 800,000 odmt/yr corresponding to a large ethanol plant.

## Administrative and ecological boundaries.

### FMU map

Many forest products companies operate in Alberta, and have different agreements
with the government with respect to how the forest is managed.  Some companies
have harvesting rights to just the needleleaf species, others have rights to
just the broadleaf species, and some have rights to both needleleaf and
broadleaf.

Utilization standards in Alberta are specified on the basis of stump height,
minimum stump diameter (outside bark), and minimum log length to a specified top
diameter (inside bark).  Utilization standards vary by FMU.  Stump heights used
in Alberta are 15 and 30 cm; stump diameters are 13 or 15 cm; top diameters are
7, 10, 11, or 12 cm; and minimum log lengths are 2.44, 2.66, 2.67, 3.55, 3.76, 
or 4.88 m.  These utilization standards are necessary to relate volumes 
calculated to the 13/7 standard for the National Forest Inventory to the harvest
volumes reported by the companies.  Perhaps most importantly, the minimum top
diameter will have an effect on the volume of residue left in the forest 
corresponding to the tops of trees.

```{r FMU, echo=FALSE, fig.cap="Location of Forest Management Units in Alberta showing Green and White Areas.", out.width = '80%'}
knitr::include_graphics("../img/FMU.png")
```

### Terrestrial ecozone map

The NFIS' individual tree biomass calculator @NFIS2016 uses province, ecozone,species, dbh, and height as input variables, so ecozone is important.  It is based on 
equations developed by Lambert
*et al.* (2005) @Lambert2005.

```{r ecozones, echo=FALSE, fig.cap="Terrestrial ecozones of Canada within Alberta.", out.width = '80%'}
knitr::include_graphics("../img/ecozones.png")
```
Example output from the biomass calculator.

```{r biomass calculator example, echo=FALSE, fig.cap="Biomass calculator example.", out.width = '80%'}
knitr::include_graphics("../img/biomass_calc_example.png")
```

## Spatial variability in residual biomass production

There is considerable spatial variability in the amount of timber harvest
residues available for biomass collection.  The variability relates to the
distribution of harvested species and age classes across the landscape, to the
timber harvesting practices of different operators, and to other factors.
We used the 250 m resolution raster maps of Canada's forest attributes for 2011
from Natural Resources Canada @Beaudoin2017 to retrieve information on species
composition, age classes, and merchantable volume.  The data were clipped to the
Alberta boundary.  Merchantable volumes for Alberta in this dataset were
compiled to a 13/7 utilization standard, meaning that stump height was assumed
to be 30 cm, that stump diameter (outside bark) needed to be at least 13 cm, and
that the length of the log to a 7 cm inside bark diameter must be at least 4.88
m for a log to be merchantable.  This is one of the utilization standards used
in Alberta and was used for Canada's National Forest Inventory (NFI henceforth)
@Boudewyn2007.

The first step was to identify cells that contained merchantable forest
inventory. We selected cells that contained at least 50 m^3^/ha of merchantable
volume, as that corresponds with merchantability limits specified in Alberta
@Alberta2016. We filtered out cells with more than 50% crown closure of larch
(*Larix spp.*) and birch (*Betula spp.*) as those genera are rarely harvested at
a commercial scale in Alberta.  We filtered out cells with stand age less than
80 years as it is unusual to harvest stands younger than that in Alberta.
This filtered raster was "unioned" with the township coverage, and an average of
the merchantable volume in the merchantable cells was calculated.  This volume
was adjusted upwards by the ratio of the average volume per hectare harvested
from the West Fraser (Hinton) Forest Management Agreement Area (FMA) to the
average merchantable volume for each township calculated from the filtered
raster and township coverages.  The multiplicative volume adjustment factor was
calculated as `r sprintf("%5.2f",merch_vol_adj_factor)`.

**A little more explanation is needed..**

Something about needleleaf and broadleaf trees.  Most common species harvested
are white spruce (_Picea glauca_), lodgepole pine (*Pinus contorta* var.
*latifolia*), trembling aspen (*Populus tremuloides*), and less commonly jack
pine (*Pinus banksiana*), black spruce (*Picea mariana*), balsam poplar
(*Populus balsamifera*), subalpine fir (*Abies lasiocarpa*), balsam fir (*Abies
balsamea*), and Douglas-fir (*Pseudotsuga menziesii* var. *glauca*).

## Service areas
** A nice picture demonstrating dendritic nature of road network **
* The provincial road network was sorted into four classes:
* primary provincial highways
* other paved roads
* two-lane gravel roads
* one-lane gravel or dirt roads
* Average speeds calculated using harmonic mean:
$s_h= \left(\frac{s_u^{-1} + s_l^{-1}}{2}\right)^{-1}$
* Speeds by road class, loaded and unloaded:
* primary provincial highways: loaded 80, unloaded 95, h-mean 85
* other paved roads: loaded 65, unloaded 80, h-mean 70
* two-lane gravel roads: loaded 50, unloaded 70, h-mean 60
* one-lane roads: loaded 20, unloaded 30, h-mean 25
* Rasterized network areas were generated using the QGIS plugin, QNEAT3
(Iso-Area as Interpolation) @Raffler2018 .
* QGIS zonal statistics were used to calculate the median network travel time in
each township.

# Theory/Calculation
The GIS analysis for this paper was done using QGIS @QGISorg2021.  Data
summaries and statistical analysis were done using R @R2021 using RStudio
@RStudioTeam2021, and tidyverse @Wickham2019.
So here we set the constants.
Following @Roser2013, the pre-piling cost was set to `r
sprintf("%5.2f",prepiling_cost)` \$/odmt, the comminution (*grinding ?*) cost
was set to `r sprintf("%5.2f",comminution_cost)`  \$/odmt, and other costs were
set to  `r sprintf("%5.2f",other_cost)` \$/odmt.  The total "constant" costs are
`r sprintf("%5.2f",constant_cost)` \$/odmt.
Using *blah blah blah* , the truck rate was set to `r
sprintf("%5.2f",truck_rate)` \$/odmt/h.
A 100 m³ trailer can haul `r odt_trailer` odmt of chipped residue.

@@

harv <- subset(t, harvest_ha > 0)
#head(harv)
#summary(harv$harvest_ha)
#summary(garv$harvest_vol) 
#summary(harv$adj_vol)

@@

#summary(harv2$harvest_vol)
#harv2

harv3 <- read.csv(file = '../data/ABHarvestVol.csv') %>% filter(year >=1990)
#harv3

@@

# ggplot(data=harv4, aes(x=year)) +
#   geom_line(aes(y = harvest_vol), colour="darkred") +
#   geom_line(aes(y = harvvol), colour="steelblue") +
#   labs(x = "year") + labs(y = "harvest volume (million m3)")  +
#   scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6)) 
# 
# #2009 is the median harvest volume year for Alberta in 1990-2015
# 
# ggplot(data=harv2, aes(x=year,y=harvest_vol)) +
#   labs(x = "year") +
#   labs(y = "harvest volume (million m3)") +
#   geom_line()  +
#   scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6)) 
# +
#   scale_colour_viridis(discrete = TRUE) +
#   theme(legend.position = "none") +
#   facet_grid(cols=vars(location)) +
#   scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))

@@

#head(d)

big_joined <- inner_join(d,harv, by = "TRM")
#head(big_joined)
#tail(big_joined)

@@

```
<!-- As a check, we obtained harvest levels for the period 1990-2018 from the National -->
<!-- Forestry Database @ccfm2020 .  Note that in 1998, industrial salvage volumes of 369,791 m³ of  -->
<!-- softwood and 264,647 m³ are not included in the harvest levels shown in the graph -->

<!-- ```{r pressure, echo=FALSE, fig.cap="NFDP harvest levels", out.width = '100%'} -->
<!-- knitr::include_graphics("../img/alberta_harvest.png") -->
<!-- ``` -->

@@

# Reference list
